Factors determining how catchments respond to forest cover change. Re-analysing global data sets.

### 2022-03-08 paper/Forest_and_Water/Forest_and_Water.tex
addressing WBs comments

  \title{Factors determining how catchments respond to forest cover change. Re-analysing global data sets.}
    \author[DARE, The University of Sydney]{R. Willem Vervoort\corref{1}}
   \ead{willem.vervoort@sydney.edu.au} 
    \author[INIA]{Eliana Nervi}
   \ead{eliananervi@gmail.com} 
    \author[IMFIA]{Jimena Alonso}
   \ead{jalonso@fing.edu.uy} 
      \address[DARE]{ARC Training Centre Data Analytics for Resources and the Environment}
    \address[The University of Sydney]{School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia}
    \address[INIA]{Project Manager, FPTA 357, Instituto Nacional de Investigacion Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay}
    \address[IMFIA]{Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow. Deforestation causes a 32\% greater change in flow compared to reforestation. Area of the catchment only affects the change in streamflow after log transformation, given the wide variety in the data from small scale paired watershed studies. Smaller studies dominate the database with 42\% of the data \textless{} 1 km2 and 65\% of the data \textless{} 10 km2. Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies. Despite these findings, overall explained variance (38\%) is low due the quality of the inputs and additional unknown confounding factors.
  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow. Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow. However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation. The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow? Generalised additive models were used to run flexible regressions including multiple variables.
  \end{abstract}

@@

For the purpose of this paper, \emph{watershed} and \emph{catchment} are interchangeable terms. Many of the US studies use \emph{watershed}, while European and Australian studies use \emph{catchment}. In particular, we retained the term ``paired watershed studies'' as this is the most common terminology, but further mostly use the term catchment.

@@

The second study (Filoso et al., 2017) was a systematic review which classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis. Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors. For a subset of quantitative data (37) they showed a log-linear relationship between decreasing catchment size and an increasing decline in streamflow. In addition, they identified that studies with shorter periods of data collection resulted in larger declines in streamflow.

A final summary paper that includes much of the same data as Zhang et al. (2017) and Filoso et al. (2017) is Zhou et al. (2015), which has one author in common with Zhang et al. (2017). However, this paper aims to explain the variation in the data using the Fuh model, and in particular aims to link the variation in the observed data to variations in the exponent \emph{m} in the model. A key observation is that in drier environments, the effects of deforestation are much greater than in wetter environments, which is also suggested by Figure 4 in Zhang et al. (2017).

Encouraged by the work presented by Zhang et al. (2017), Filoso et al. (2017) and Zhou et al. (2015) and the large database of studies presented by these authors, we believe more can be done to add to this important discussion. In this paper, the aim extend the analysis of the collected data and to expand and combine the data sets.

@@

While Zhang et al. (2017) use the dryness index in their analysis, and calculate the variable climate type from this index, the potential or reference evapotranspiration was not originally included as part of the published data set. In this paper, we used only the dryness index and did not use the climate type as a variable (as they are interchangeable).
We combined the tables for small catchments (\textless{} 1000 km\textsuperscript{2}) and large catchments (\textgreater= 1000 km\textsuperscript{2}) in our analysis.

@@

To enhance the existing data set, this study added additional variables and cross-checked the studies with the data set from Filoso et al. (2017). In particular, we focussed on the 37 data points included in the quantitative analysis in Filoso et al. (2017).

In addition, additional variables added were the latitude and longitude for the center of the catchment as an approximation of its spatial location. Using this information reference evapotranspiration (E\textsubscript{0}) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET\textsubscript{0}) Climate Databasev2 (Trabucco and Zomer, 2018), if a value of E\textsubscript{0} was not available from the original papers. For large catchments, this value, similar to annual average rainfall, is only an approximation of the climate at the location.

@@

The length of the study can be a variable influencing the change in flow (Filoso et al., 2017; e.g. Jackson et al., 2005), as for example, more mature plantations are thought to have smaller impacts on flow or regrowth might follow a ``Kuczera curve'' (Kuczera, 1987). Therefore, the length of the study calculate as the difference between the starting data and completion date of the different studies was extracted from the references provided by Zhang et al. (2017). The length of the study was already included in the data from Filoso et al. (2017), but these were checked against the original publications.

Several additional data points from catchment studies were extracted from Zhang et al. (2011), Zhao et al. (2010), Borg et al. (1988), Thornton et al. (2007), Zhou et al. (2010), Rodriguez et al. (2010), Ruprecht et al. (1991) and Peña-Arancibia et al. (2012), and these were checked against the existing studies to prevent overlap. In the citation column in the data set, in general the main reference for the calculated change in streamflow was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.~Table 6 in Zhang et al. (2011)).
We also removed one data point from the analysis, which corresponds to catchment \#1 (Amazon) in Zhang et al. (2017). This is because the cited reference (Roche, 1981) only relates to 1 and 1.5 ha paired catchment studies in French Guyana, and in which the actual change in forest cover is not recorded. Furthermore, the change in flow for catchment \#76 was corrected from 600\% to 157\% after review of the original publication (Baker Jr., 1984). Finally, on review of all the data in Zhang et al. (2017) and Filoso et al. (2017), 29 potential duplicates were identified and flagged in the data.

The final column in the improved data set is a ``notes'' column, which is not further used in the analysis, but gives context to some of the data for future research and highlights some of the discrepancies that we found between the original papers and the data in the tables from Zhang et al. (2017).

@@

Here \(X_i\) are factorial variables, while \(Z_i\) are continuous variables. The model assumes no direct interactions and all variables are additive. The changes in forest cover contain both positive (forestation) and negative values (deforestation). In Zhang et al. (2017), these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect if removing forest cover was the same as an equivalent reforestation. However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover. Therefore all the change in forest cover data is converted to positive values, and an additional column (\(sign_{forest cover}\)) is added that indicates whether it was a forest cover increase or decrease.

A further assumption in the model is that all continuous variables \(Z_i\) (such as annual precipitation (Pa)) can have a linear or non-linear relationship with \(\Delta Qf \%\). This means that a smooth function \(s()\) is applied to the \(Z_i\) variables. For the smoothing function we applied thin plate regression splines with an additional shrinkage penalty which means the terms can be shrunk to 0 if not significant (Wood, 2006).

For the model in equation 2, we initially only used the data from Zhang et al. (2017) to make sure that the additional catchments added to the data set did not influence the results. Subsequently the analysis was repeated and the additionally identified catchments were added.

More generally the results were analysed to identify:\\
1. the significance of the different variables\\
2. the direction of the categorical or shape of the smooth variables

@@

The overall dataset contains 330 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data. In contrast, the original dataset from Zhang et al. (2017) contained 312 catchments and the Filoso et al. (2017) study used 37 catchments (Table S2 in Filoso et al. (2017)). The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and \emph{Area km\^{}2}. The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution. Note also the large number of studies with 100\% forest cover reduction. Smaller catchments dominate the database with 42\% of the data from catchments \textless{} 1 km\textsuperscript{2} and 65\% of the data for catchments \textless{} 10 km\textsuperscript{2} (Figure 1).

\begin{figure}
\includegraphics[width=0.9\linewidth]{./DataExploration} \caption{Overview of the distribution of the data set for five of the included variables.}\label{fig:datagraphs}
\end{figure}

In addition, this shows that for the data related to forest decreases, there is almost always a positive flow change (Figure 2). In other words, flow almost always increased. However, for increases in forest cover, this is not the case, and flow can both increase and decrease. However in both cases the variability in the reported change in flow increases with the increase in forest cover change.

\begin{figure}
\includegraphics[width=0.9\linewidth]{Increase_decrease} \caption{Changes in flow as a function of increases and decreases in forest cover}\label{fig:increasedecrease}
\end{figure}

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
\caption{\label{tab:tabmodel1} Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change. The first three rows relate to the model using the original data base from Zhang et al.~(2017). The bottom three rows are the results of the model including the new data.}\tabularnewline
\toprule
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule
\endhead
\textbf{(Intercept)} & 8.65 & 5.56 & 1.56 & 0.12 \\
\textbf{DeltaF\_perc\_pos} & 0.45 & 0.09 & 5.26 & 0 \\
\textbf{Forest\_Signincrease} & -29.17 & 5.79 & -5.04 & 0 \\
\textbf{(Intercept)} & 10.76 & 5.34 & 2.02 & 0.04 \\
\textbf{DeltaF\_perc\_pos} & 0.41 & 0.08 & 5.04 & 0 \\
\textbf{Forest\_Signincrease} & -38.03 & 5.26 & -7.23 & 0 \\
\bottomrule
\end{longtable}

@@

Including the data from some of the newly identified studies indicates that this mainly strengthens the difference between the forest cover increases and decreases (Table \ref{tab:tabmodel1}), and the result indicate a reduction in the mean decrease in flow as a result of forest cover change if the new data is included.

@@

As indicated, an initial hypothesis relates to whether there is a strong spatial global gradient as captured by latitude and longitude. These data were added for the different studies, mostly by using the data reported by the authors, but in some cases approximating the location of the centre of the catchment using Google Maps. In the dataset, an additional column is added to indicate the source of the location data. As the global map (Figure \ref{fig:globalmap}) shows, the distribution of case study catchments covers multiple continents and shows some distinct clustering in parts of the world. Of interest is whether the spatial clustering also indicates a difference in response to forest cover change:

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
\caption{\label{tab:out-model3} Results of the model including Latitude and Longitude including new data}\tabularnewline
\toprule
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule
\endhead
\textbf{(Intercept)} & 10.99 & 5.75 & 1.91 & 0.06 \\
\textbf{DeltaF\_perc\_pos} & 0.42 & 0.08 & 5.13 & 0 \\
\textbf{Forest\_Signincrease} & -38.99 & 5.54 & -7.04 & 0 \\
\textbf{Latitude} & -0.01 & 0.09 & -0.13 & 0.9 \\
\textbf{Longitude} & 0.01 & 0.03 & 0.26 & 0.79 \\
\bottomrule
\end{longtable}

There appears to be no significant gradient in either latitude or longitude (Table \ref{tab:out-model3}), suggesting that the distribution of the catchments across the globe has little influence on the overall result. The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.23 suggesting further factors influencing the change in streamflow that are currently not included in the model.

@@

Climate, and in particular evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlighted by both Zhang et al. (2017) and Jackson et al. (2005). Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics). Initially, we tested models using annual average precipitation (\emph{Pa (mm)}), but because of the interactions between precipitation, evapotranspiraton and the dryness index, we concentrated on the dryness index as the key variable. Given that Latitude and Longitude were not significant, we dropped these from the model.

@@

The major hypothesis to test is the effect of area on the change in flow, following the analysis by Zhang et al. (2017) and Filoso et al. (2017). Given the highly skewed distribution of the catchment areas (Figure \ref{fig:datagraphs}), a log base 10 transformation was applied to the variable \emph{Area (km\textsuperscript{2})}.

@@

The work by Filoso et al. (2017) and earlier by Jackson et al. (2005) has indicated that the length of the study might influence the response. This links to the idea from Kuczera (1987) that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also Jackson et al. (2005)). In addition to adding \emph{length} (being the difference between the reported start date and end date of data collection in the specific study) as a variable, three other continuous variables (\emph{Dryness, Area, From}) were considered non-linear in this model. As a result a shrinkage smoothing spline (Wood, 2006) was applied to these variables. \emph{From} represents the starting date of the data collection.

\begin{align}
\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \\ & s(From) + \varepsilon \label{eq:eq7}
\end{align}

>>>>

  \title{Factors determining how catchments respond to forest cover change. Re-analysing global data sets.}
    \author[DARE]{R. Willem Vervoort\corref{1}}
   \ead{willem.vervoort@sydney.edu.au} 
    \author[INIA]{Eliana Nervi}
   \ead{eliananervi@gmail.com} 
    \author[IMFIA]{Jimena Alonso}
   \ead{jalonso@fing.edu.uy} 
      \address[DARE]{ARC Training Centre Data Analytics for Resources and Environments \& Sydney Institute of Agriculture, School of Life and Environmental Sciences.}
    \address[The University of Sydney]{The University of Sydney, Sydney, NSW 2006, Australia}
    \address[INIA]{Project Manager, FPTA 357, Instituto Nacional de Investigacion Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay}
    \address[IMFIA]{Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow. Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow. However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation. The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow? Generalised additive models were used to run flexible regressions including multiple variables.
  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow. Removal of forest cover causes a 32\% greater change in flow relative to increasing forest cover. Area of the catchment only affects the change in streamflow after log transformation, due to high skew in the data. Smaller catchment dominate the database with 42\% of the data \textless{} 1 km\textsuperscript{2} and 65\% of the data \textless{} 10 km\textsuperscript{2}. Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies. Despite these findings, overall explained variance (38\%) of the regression model is low due the quality of the inputs and additional unknown confounding factors.
  \end{abstract}

@@

For the purpose of this paper, \emph{watershed} and \emph{catchment} are interchangeable terms. Many of the US studies use \emph{watershed}, while European and Australian studies use \emph{catchment}. In particular, we retained the term ``paired watershed studies'' and ``quasi-paired watershed studies'' as this is the most common terminology, but further mostly use the term catchment.

@@

The second study (Filoso et al., 2017) is a systematic review of reforestation studies (only studies in which forest cover increased). This study classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis. Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors. For a subset of the data (37 data points) they also indicated decreasing impacts of reforestation with increasing catchment size (agreeing with Zhang et al. (2017)), but they did not identify a distinct threshold and fitted a log-linear relationship. In addition, they identified that studies with shorter periods of data collection resulted in larger declines in streamflow.

A final summary paper that includes much of the same data as Zhang et al. (2017) and Filoso et al. (2017) is Zhou et al. (2015), which has one author in common with Zhang et al. (2017). However, this paper aims to explain the variation in the data using the elasticity approach in the Fuh model. In particular, it aims to link the variation in the observed data to variations in the exponent \emph{m} in the Fuh model. A key observation is that in drier environments, the effects of removing forest cover are much greater than in wetter environments, which is also suggested by Figure 4 in Zhang et al. (2017).

Encouraged by the work from Zhang et al. (2017), Filoso et al. (2017) and Zhou et al. (2015) and the large database of studies presented by these authors, we believe more can be done to add to this important discussion. In this paper, the aim is to extend the analysis of the collected data and to expand and combine the data sets.

@@

While Zhang et al. (2017) use the dryness index in their analysis, and calculate the variable climate type from this index, the potential or reference evapotranspiration was not originally included as part of the published data set. In addition, dryness might mask areas where high rainfall (with potentially higher intensity rainfall) dominates the impact of high ET. In other words, high rainfall can possibly point to more infiltration excess runoff, which might be less impacted by catchment wetness condition (determined by cumulative ET). In this paper, we do include the dryness index but did not use the climate type as a variable (as they are interchangeable).
We combined the tables for small catchments (\textless{} 1000 km\textsuperscript{2}) and large catchments (\textgreater= 1000 km\textsuperscript{2}) from Zhang et al. (2017) in our analysis.

@@

To enhance the existing data set, this study added additional variables and cross-checked the studies with the data set from Filoso et al. (2017). In particular, we focussed on including the 37 data points related to the quantitative analysis in Filoso et al. (2017).

In addition, latitude and longitude for the center of the catchment as an approximation of its spatial location. These data were added for the different studies, mostly by using the data reported by the authors, but in some cases approximating the location of the centre of the catchment using Google Maps\textsuperscript{TM}. In the dataset, an additional column has been added to indicate the source of the location data.

Climate more generally, and in particular the ratio of rainfall and evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlighted by both Zhang et al. (2017) and Jackson et al. (2005). Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics). Using the location information reference evapotranspiration (E\textsubscript{0}) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET\textsubscript{0}) Climate Databasev2 (Trabucco and Zomer, 2018), if a value of E\textsubscript{0} was not available from the original papers. For large catchments, this value (and the associated coordinates), similar to annual average rainfall, is only an approximation of the climate at the location.

@@

The length of the study can be a variable influencing the change in flow (Filoso et al., 2017; e.g. Jackson et al., 2005), as for example, more mature plantations are thought to have smaller impacts on flow or regrowth might follow a ``Kuczera curve'' (Kuczera, 1987). It is not clear if this is an effect of increased water use in growth (Vertessy et al., 2001) or due to changes in interception (Stoof et al., 2012). Therefore, the length of the study calculate as the difference between the starting data and completion date of the different studies was extracted from the references provided by Zhang et al. (2017). The length of the study was already included in the data from Filoso et al. (2017), but these were checked against the original publications.

Several additional data points from catchment studies were extracted from Zhang et al. (2011), Zhao et al. (2010), Borg et al. (1988), Thornton et al. (2007), Zhou et al. (2010), Rodriguez et al. (2010), Ruprecht et al. (1991) and Peña-Arancibia et al. (2012), and these were checked against the existing studies to prevent overlap. In the citation column in the accompanying data set, the main reference for the calculated change in streamflow was generally used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e.~Table 6 in Zhang et al. (2011)).
We also removed one data point from the analysis, which corresponds to catchment \#1 (Amazon) in Zhang et al. (2017). This is because the cited reference (Roche, 1981) only relates to 1 and 1.5 ha paired catchment studies in French Guyana, and in which the actual change in forest cover is not recorded. Furthermore, the change in flow for catchment \#76 was corrected from 600\% to 157\% after review of the original publication (Baker Jr., 1984). Finally, on review of all the data in Zhang et al. (2017) and Filoso et al. (2017), 29 potential duplicates were identified and flagged in the data, and not used in the analysis.

The final column in the improved data set is a ``notes'' column, which we added, but is not further used in the analysis. It gives context to some of the data for future research and highlights some of the discrepancies that we found between the original papers and the data in the tables from Zhang et al. (2017). This will allow future research to further sructinise our input for errors.

@@

Here \(X_i\) are factorial variables, while \(Z_i\) are continuous variables. The model intially assumes no direct interactions and all variables are additive. We will comment in this assumption in the discussion. The changes in forest cover contain both positive (forestation) and negative values (deforestation). In Zhang et al. (2017), these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover. However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover. Therefore all the change in forest cover data was converted to positive values, and an additional factorial column (\(sign_{forest cover}\)) is included indicating whether it was a forest cover increase or decrease.

A further assumption in the model is that all continuous variables \(Z_i\) (such as annual precipitation (Pa)) can have either a linear or a non-linear relationship with \(\Delta Qf \%\). This means that a smooth function \(s()\) can be applied to the \(Z_i\) variables. For the smoothing function we applied thin plate regression splines with an additional shrinkage penalty. The result of this approach is that for high enough smoothing parameters (i.e.~if the data is very ``wiggly'') the smooth term can be shrunk to 0 and thus will be no longer significant (Wood, 2006). This is done because a highly flexible smooth term could always fit the data, but would not necessarily indicate a relevant relationship. In other words, the approach balances finding a smooth non-linear relationship for the variable against overfitting the data.

The over arching test focuses on the change streamflow as a result of a change in forest cover being influenced by three major additional factors (as indicated by the previous research: Zhang et al. (2017); Filoso et al. (2017); Zhou et al. (2015)): climate, size of catchment and length of study. Therefore, even if these variables are insignificant in any of the applied models, we retained variables representing these three factors.

As an initial approach we only used the data from Zhang et al. (2017) to make sure that the additional catchments added to the data set did not influence the results (this is discussed in the results). Subsequently the analysis was repeated and the additionally identified catchments were added.

@@

The overall dataset contains 327 observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data. In contrast, the original dataset from Zhang et al. (2017) contained 312 catchments and the Filoso et al. (2017) study used 37 catchments (Table S2 in Filoso et al. (2017)). The current number of catchments is the result of the removal of duplicates and our modifications and additions. The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and \emph{Area km\textsuperscript{2}}. The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution. Note also the large number of studies with 100\% forest cover reduction. Clearly visible is also that smaller catchments dominate the database with 42\% of the data from catchments \textless{} 1 km\textsuperscript{2} and 65\% of the data for catchments \textless{} 10 km\textsuperscript{2} (Figure 1).

\begin{figure}
\includegraphics[width=0.9\linewidth]{./DataExploration} \caption{Overview of the distribution of the data set for five of the included variables. Note that the first panel (showing the distribution of the catchment areas) indicates the distribution of the \emph{log\_10\_} transformed Area (in km\textsuperscript{2}).}\label{fig:datagraphs}
\end{figure}

Analysing this in more detail, the data related to forest decreases, indicate almost always a positive flow change (Figure 2). In other words, flow almost always increased. However, for increases in forest cover, this is not the case, and flow can both increase and decrease. However in both cases the variability in the reported change in flow increases with the increase in forest cover change.

\begin{figure}
\includegraphics[width=0.9\linewidth]{Increase_decrease} \caption{Changes in flow as a function of decreases (top) and increases (bottom) in forest cover}\label{fig:increasedecrease}
\end{figure}

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
\caption{\label{tab:tabmodel1} Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change. The first three rows relate to the model using the original data base from Zhang et al.~(2017). The bottom three rows are the results of the model including the new data. Clearly there is no major change arising from the additional data.}\tabularnewline
\toprule
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule
\endhead
\textbf{(Intercept)} & 8.65 & 5.56 & 1.56 & 0.12 \\
\textbf{DeltaF\_perc\_pos} & 0.45 & 0.09 & 5.26 & 0 \\
\textbf{Forest\_SignIncrease} & -29.17 & 5.79 & -5.04 & 0 \\
\textbf{(Intercept)} & 10.12 & 5.32 & 1.9 & 0.06 \\
\textbf{DeltaF\_perc\_pos} & 0.42 & 0.08 & 5.24 & 0 \\
\textbf{Forest\_SignIncrease} & -36.8 & 5.26 & -6.99 & 0 \\
\bottomrule
\end{longtable}

@@

Including the data from some of the newly identified studies indicates that this mainly strengthens the difference between the forest cover increases and decreases (Table \ref{tab:tabmodel1}), and the result indicate a reduction in the mean decrease in flow as a result of forest cover change if the new data is included. Adding the new data does not change the outcome much (apart from the magnitudes of the coefficients), which is expected as the number of added catchments is small relative to the total Zhang et al. (2017) data set. But this also means that our re-analysis of the data can be directly compared to the original study.

@@

Latitude and longitude might reveal strong spatial clustering of the studies, or latitude and longitude might indicate strong climate gradients. As the global map (Figure \ref{fig:globalmap}) shows, the distribution of case study catchments covers multiple continents and shows some distinct clustering in parts of the world. Of interest is whether the spatial clustering also indicates a difference in response to forest cover change:

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.36}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.18}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.14}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
\caption{\label{tab:out-model3} Results of the model based on the complete dataset and including Latitude and Longitude}\tabularnewline
\toprule
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{\linewidth}\centering
~
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Std. Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
t value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Pr(\textgreater\textbar t\textbar)
\end{minipage} \\
\midrule
\endhead
\textbf{(Intercept)} & 14.74 & 5.35 & 2.76 & 0.01 \\
\textbf{DeltaF\_perc\_pos} & 0.44 & 0.08 & 5.5 & 0 \\
\textbf{Forest\_SignIncrease} & -35.72 & 5.47 & -6.53 & 0 \\
\textbf{Latitude} & -0.18 & 0.05 & -3.45 & 0 \\
\textbf{Longitude} & -0.01 & 0.01 & -1.89 & 0.06 \\
\bottomrule
\end{longtable}

There appears to be no significant gradient in either latitude or longitude (Table \ref{tab:out-model3}), suggesting that the distribution of the catchments across the globe has little influence. The total explaining power of the model is still low with an adjusted \emph{r\textsuperscript{2}} of 0.27 suggesting further factors influencing the change in streamflow that are currently not included in the model.

@@

While latitude and longitude might hint at climatic gradients (for example a change in response related to tropical or sub tropical belts), annual rainfall and potential evapotranspiration might give a better indication.
Potential evapotranspiration (\emph{E0}) by itself was not significant in the. Initially, we also tested models using only the annual average precipitation (\emph{Pa (mm)}), but interactions between precipitation and evapotranspiraton might be captured by the dryness index. Both dryness index and \emph{Pa (mm)} were initially analysed as a key variables, but these indicated that these two variables were essentially interchangeable. As a result only the dryness index was retained as a climate indicator to align with the earlier work by Zhang et al. (2017). Given that Latitude and Longitude were not significant, we dropped these from the model.

@@

The second major variable is the effect of area on the change in flow, following the analysis by Zhang et al. (2017) and Filoso et al. (2017). Given the highly skewed distribution of the catchment areas (Figure \ref{fig:datagraphs}), a log base 10 transformation was applied to the variable \emph{Area (km\textsuperscript{2})}.

@@

The work by Filoso et al. (2017) and earlier by Jackson et al. (2005) has indicated that the length of the study might influence the response. This links to the idea from Kuczera (1987) that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also Jackson et al. (2005)). In addition to adding \emph{length} (being the difference between the reported start date and end date of data collection in the specific study) as a variable, two other continuous variables (\emph{Dryness} and \emph{Area}) were considered non-linear. As a result a shrinkage smoothing spline (Wood, 2006) was applied to these variables.

\begin{align}
\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \varepsilon \label{eq:eq7}
\end{align}
